---
title: "Data cleaning"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Research Question

My interest in this area came from multiple new information I've received recently. Since I have been taking Asian studies courses and came to learn how unprivileged some countries are, I have been wanting to learn more about them. Another cause of the interest is I have been learning more about woman's birth stories and the amount of preparation and care they need to deliver a baby (or even more) safely. Recently, this topic has been on my mind so I thought it would be a great opportunity to do more research on this area.

## Background

I looked up the World Health Organization website regarding the topic of maternaity morality rate and possible factors to affect it. According to the article, factors including but not limited to: severe bleeding, infection, high blood pressure, complications from delivery, and unsafe abortion, could affect the rate of maternity mortality.
World Health Organization. (2019, September 19). Maternal mortality. 

## Data Source

The original dataset has multiple tabs included. We're only focusing on the ICPD_health tab in this project.
In the dataset, there are multiple columns with various factors included. We're only keeping a few to conduct the research. From the WHO website I visited, the factors included in the dataset that are related to the factors mentioned are: Percentage of births attended by skilled health personnel and new HIV infections. 
```{r, include=FALSE}
library(readxl)
SWOP_Data_2021 <- read_excel("/users/sunnywang/Downloads/SWOP-Data-2021.xlsx", 
    sheet = "ICPD_Health_2021", skip = 1, n_max = 210)
summary(SWOP_Data_2021)
#We are pulling out the columns we need to produce another dataset
maternal_mortality_data <- select(SWOP_Data_2021, 'Countries, territories, other areas', 'Maternal mortality ratio (MMR) (deaths per 100,000 live births)a', 'Births attended by skilled health personnel, percent', 'Number of new HIV infections, all ages, per 1,000\r\nuninfected population')

#Delete rows with NA values inside
maternal_mortality_data <- maternal_mortality_data %>% mutate_all(funs(as.character(.)))

#There were numbers that weren't entered in correct format that we need to fix
maternal_mortality_data$'Number of new HIV infections, all ages, per 1,000\r\nuninfected population'[maternal_mortality_data$'Number of new HIV infections, all ages, per 1,000\r\nuninfected population' == "7.0000000000000007E-2"] <- "0.07" 

#There are still empty values which are denoted as '-' that we need to remove from the data
maternal_mortality_data$'Number of new HIV infections, all ages, per 1,000\r\nuninfected population'[maternal_mortality_data$'Number of new HIV infections, all ages, per 1,000\r\nuninfected population' == "-"] <- NA 
maternal_mortality_data$'Countries, territories, other areas'[maternal_mortality_data$'Countries, territories, other areas' == "-"] <- NA 
maternal_mortality_data$'Maternal mortality ratio (MMR) (deaths per 100,000 live births)a'[maternal_mortality_data$'Maternal mortality ratio (MMR) (deaths per 100,000 live births)a' == "-"] <- NA 

maternal_mortality_data$'Births attended by skilled health personnel, percent'[maternal_mortality_data$'Births attended by skilled health personnel, percent' == "-"] <- NA 

#Dropped all values that are NA
maternal_mortality_data <- drop_na(maternal_mortality_data, 'Countries, territories, other areas', 'Maternal mortality ratio (MMR) (deaths per 100,000 live births)a','Births attended by skilled health personnel, percent', 'Number of new HIV infections, all ages, per 1,000\r\nuninfected population')

#Rename columns since the original names are too lengthy
maternal_mortality_data <- rename(maternal_mortality_data, 'Locations'= 'Countries, territories, other areas', 'Maternal Mortality Rate/100,000 births' = 'Maternal mortality ratio (MMR) (deaths per 100,000 live births)a', 'New HIV Infections/1000 people' = 'Number of new HIV infections, all ages, per 1,000\r\nuninfected population', 'Birth Attended by Professionals (%)' = 'Births attended by skilled health personnel, percent')
maternal_mortality_data
class(maternal_mortality_data$`Maternal Mortality Rate/100,000 births`)
class(maternal_mortality_data$`New HIV Infections/1000 people`)
class(maternal_mortality_data$`Birth Attended by Professionals (%)`)
```
```{r, echo=FALSE}
#I had to change the variable type from discrete to continuous in order to plot them
ggplot(maternal_mortality_data, aes(as.numeric(as.character(`New HIV Infections/1000 people`)))) + geom_histogram(bins = 30) + labs(x = 'New HIV Infections per 1000 people', title = '(Figure 1) Histogram of New HIV Infections per 1000 people')

ggplot(maternal_mortality_data, aes(as.numeric(as.character(`Birth Attended by Professionals (%)`)))) + geom_histogram(bins = 30) + labs(x = 'Birth Attended by Professionals (%)', title = '(Figure 2) Histogram of Birth Attended by Professionals (%)')

ggplot(maternal_mortality_data, aes(as.numeric(as.character(`Maternal Mortality Rate/100,000 births`)))) + geom_histogram(bins = 30) + labs(x = 'Maternal Mortality Rate/100,000 births', title = '(Figure 3) Histogram of Birth Attended by Professionals (%)')
```
In Figure 1 we can see the graph is skewed to the right with possible outliers on the right side of the graph. This figure has a downward sloping trend. In Figure 2 we can see the graph is skewed to the left with possible outliers around 20% of birth attended by professionals. In figure 3, the graph is again skewed to the right with possible outlier on the right side of the graph, around 1200 mortality at birth per 100,000 births. We also want to note that locations without the presence of all three data we need are deleted. Therefore, the skewedness in these graphs could be caused by the absence of those data.
```{r, echo=FALSE}
ggplot(maternal_mortality_data, aes(as.numeric(as.character(`Birth Attended by Professionals (%)`)),as.numeric(as.character(`Maternal Mortality Rate/100,000 births`)))) + geom_point() + labs(x = 'Birth Attended by Professionals (%)', y = 'Maternal Mortality Rate/100,000 births', title = '(Figure 4) Birth Attended by Professionals (%) v. Maternal Mortality Rate')

ggplot(maternal_mortality_data, aes(as.numeric(as.character(`New HIV Infections/1000 people`)),as.numeric(as.character(`Maternal Mortality Rate/100,000 births`)))) + geom_point()+ labs(x = 'New HIV Infections/1000 people', y = 'Maternal Mortality Rate/100,000 births', title = '(Figure 5) New HIV Infections/1000 people v. Maternal Mortality Rate')
```
In figure 4, there is a weak negative linear relationship between the two factors. The scatterplot clutters around 100% birth attended by professionaals, In figure 5, there's a weak linear relationship between the two factors with a clutter around 0 new HIV infections per people in relation to maternal mortality rate. 

My research question can be used using linear models because there are linear relationships (but weak) between variables. But some issues worth mentioning is the data we have is only from one year. Perha[s if we increase the number of years in the data the linear relationship would be stronger. 

## Reference
World Health Organization. *Maternal mortality* Retrieved October 22, 2021, from [https://www.who.int/news-room/fact-sheets/detail/maternal-mortality#:~:text=The%
20major%20complications%20that%20account,(pre%2Declampsia%20and%20eclampsia)](https://www.who.int/  news-room/fact-sheets/detail/maternal-mortality#:~:text=The%20major%20complications%20that%20account, (pre%2Declampsia%20and%20eclampsia)

